Model Recognition as Planning

### 2018-09-19 doc/icaps-19/model-recognition.tex
updated page 2 icaps19 paper

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a {\em classical planning problem} such that the semantics of the actions in $A$ is unknown because functions $\rho(s,a)$ and/or $\theta(s,a)$ are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite and non-empty {\em set of models} for the actions in $A$. Each model in $\mathcal{M}\in M$ defines a different function pair ($\rho$,$\theta$).
\item $\tau$ is an {\em observation} of the execution of a solution plan $\pi$ for $P$.
\end{itemize}
A solution to the {\em model recognition} task is the discrete probability distribution $P(\mathcal{M}|\tau)$ that expresses, for each model $\mathcal{M}\in M$, its probability of producing the observation $\tau$.

According to the {\em Bayes} rule, the probability of an hypothesis $\mathcal{H}$, provided the observation $\mathcal{O}$, is given by the expression $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In the {\em model recognition} task, hypotheses are about the possible action models $\mathcal{M}\in M$ while the given observation is the partially observed plan execution $\tau$. The $P(\mathcal{M}|\tau)$ probability distribution can then be estimated in three steps:
\begin{enumerate}
\item Computing the {\em a priori} probabilities. $P(\tau)$, that measures how surprising is the given observation and $P(\mathcal{M})$, that expresses if one model is a priori more likely than the others. 
\item Computing the conditional probability $P(\tau|\mathcal{M})$.  Our approach is to estimate this value according to the {\em amount of edits} required by the model $\mathcal{M}$ to produce a plan $\pi_\tau$ such that $\tau$ is an observation of the execution of $\pi_\tau$ on the classical planning problem $P$. The {\em edit distance} is a similarity metric that is traditionally computed over strings or graphs and that has been proved succesful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work this similarity metric refers to the edition of classical planning models. 
\item Applying the Bayes rule to obtain the normalized posterior probabilities. The $P(\mathcal{M}|\tau)$ probabilities must sum 1 for all the $\mathcal{M}\in M$.
\end{enumerate}

@@

\begin{figure}
\begin{tiny}
\begin{verbatim}
;;; Propositional encoding for inc-x(?v1 ?v2)
(pre_x-coord_v1_inc-x) (pre_next_v1_v2_inc-x) (pre_even__inc-x)
(del_x-coord_v1_inc-x) (add_x-coord_v2_inc-x)

;;; Propositional encoding for dec-x(?v1 ?v2)
(pre_x-coord_v1_dec-x) (pre_next_v2_v1_dec-x) (pre_odd___dec-x)
(del_x-coord_v1_dec-x) (add_x-coord_v2_dec-x)

;;; Propositional encoding for inc-y-odd(?v1 ?v2)
(pre_y-coord_v1_inc-y-odd) (pre_next_v1_v2_inc-y-odd) 
(pre_even__inc-y-odd)
(del_y-coord_v1_inc-y-odd) (del_odd__inc-y-odd)
(add_y-coord_v2_inc-y-odd) (add_even__inc-y-odd)

;;; Propositional encoding for inc-y-even(?v1 ?v2)
(pre_y-coord_v1_inc-y-even) (pre_next_v1_v2_inc-y-even)
(pre_even__inc-y-even)
(del_y-coord_v1_inc-y-even) (del_even__inc-y-even)
(add_y-coord_v2_inc-y-even) (add_odd__inc-y-even)

;;; Propositional encoding for dec-y-odd(?v1 ?v2)
(pre_y-coord_v1_dec-y-odd) (pre_next_v2_v1_dec-y-odd)
(pre_odd__dec-y-odd)
(del_y-coord_v1_dec-y-odd) (del_odd__dec-y-odd)
(add_y-coord_v2_dec-y-odd) (add_even__dec-y-odd)

;;; Propositional encoding for inc-y-even(v1 ?v2)
(pre_y-coord_v1_dec-y-even) (pre_next_v2_v1_dec-y-even)
(pre_even__dec-y-even)
(del_y-coord_v1_dec-y-even) (del_even__dec-y-even)
(add_y-coord_v2_dec-y-even) (add_odd__dec-y-even)
\end{verbatim}
\end{tiny}
 \caption{\small Propositional encoding for the six schema from Figure~\ref{fig:model-example}.}
\label{fig:encoding}
\end{figure}

@@

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action inc-x
  :parameters (?v1 ?v2)
  :precondition
    (and (or (not (pre_x-coord_v1_inc-x)) (x-coord ?v1))
         (or (not (pre_x-coord_v2_inc-x)) (x-coord ?v2))
         (or (not (pre_y-coord_v1_inc-x)) (x-coord ?v1))                       
         (or (not (pre_y-coord_v2_inc-x)) (x-coord ?v2))
         (or (not (pre_even__inc-x)) (even))
         (or (not (pre_odd__inc-x)) (odd)))
         (or (not (pre_next_v1_v1_inc-x)) (next ?v1 ?v1)))
         (or (not (pre_next_v1_v2_inc-x)) (next ?v1 ?v2)))
         (or (not (pre_next_v2_v1_inc-x)) (next ?v2 ?v1)))
         (or (not (pre_next_v2_v2_inc-x)) (next ?v2 ?v2)))         
    )
    :effect (and
       (when (del_x-coord_v1_inc-x) (not (x-coord ?v1)))
       (when (del_x-coord_v2_inc-x) (not (x-coord ?v2)))
       (when (del_y-coord_v1_inc-x) (not (x-coord ?v1)))
       (when (del_y-coord_v2_inc-x) (not (x-coord ?v2)))
       (when (del_even__inc-x) (not (even)))
       (when (del_odd__inc-x) (not (odd)))
       (when (del_next_v1_v1_inc-x) (not (next ?v1 ?v1)))
       (when (del_next_v1_v2_inc-x) (not (next ?v1 ?v2)))
       (when (del_next_v2_v1_inc-x) (not (next ?v2 ?v1)))
       (when (del_next_v2_v2_inc-x) (not (next ?v2 ?v2)))

       (when (add_x-coord_v1_inc-x) (x-coord ?v1))
       (when (add_x-coord_v2_inc-x) (x-coord ?v2))
       (when (add_y-coord_v1_inc-x) (x-coord ?v1))
       (when (add_y-coord_v2_inc-x) (x-coord ?v2))
       (when (add_even__inc-x) (even))
       (when (add_odd__inc-x) (odd))
       (when (add_next_v1_v1_inc-x) (next ?v1 ?v1))
       (when (add_next_v1_v2_inc-x) (next ?v1 ?v2))
       (when (add_next_v2_v1_inc-x) (next ?v2 ?v1))
       (when (add_next_v2_v2_inc-x) (next ?v2 ?v2)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Programmable version of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid.}
\label{fig:programmable}
\end{figure}

@@

For each action schema $\xi$, given the set of {\em state predicates} $\Psi$, then $F_{\xi}$ defines the subset of elements that can appear in the preconditions and effects of that action schema. For the actions schema defined in~\ref{fig:model-example} this set is the same and contains the following ten elements, {\small\tt\{x-coord($v_1$), x-coord($v_2$), y-coord($v_1$), y-coord($v_2$), odd(), even(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

\begin{enumerate}
\setcounter{enumi}{2}
\item ${\mathcal C}$, a set of {\em syntactic constraints}. These constraints include the \strips\ constraints (we assume that $\eff^-(a)\subseteq \pre(a)$, $\eff^-(a)\cap \eff^+(a)=\emptyset$ and $\pre(a)\cap \eff^+(a)=\emptyset$) and domain-specific constraints. For instace, in a {\em robot navigation} domain like the modeled in Figure~\ref{fig:model-example}, predicates {\small\tt even()} and {\small\tt odd()} are exclusive so they cannot hold a the same time.
\end{enumerate}

@@

\subsection{The $P(\mathcal{M}|\tau)$ probability for \strips\ models}
Since we assume that the dynamics of the actions $A\in P$ is specified with \strips\ action schemas, we can estimate the $P(\mathcal{M}|\tau)$ probability as follows:
\begin{enumerate}
\item Assuming {\em a priori} all action models $\mathcal{M}\in M$ are equiprobable. There are no reasons to assume that one model is {\em a priori} more likely than the others so, $P(\mathcal{M})=\frac{1}{\prod_{\xi\in\mathcal{M}} 2^{2\times|F_{\xi}|}}$.
\item Assuming that all the observations of plan executions with maximum $n$ observed actions and $m$ observed states are equiprobable then, $P(\tau)=\frac{1}{|A|^{n}\times2^{m\times|F|}}$.
\item Estimating the conditional probability $P(\tau|\mathcal{M})$ by mapping the {\em observation edit distance} into a $[0,1]$ likelihood, $1-\frac{\delta(\mathcal{M},\tau)}{\delta(\mathcal{M},*)}$.
\end{enumerate}

@@

\item Actions for {\em applying} an edited operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$ from {\em blocksworld}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
&\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{edit}\}\rhd\{\neg mode_{edit}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

\begin{figure}
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small PDDL action for applying an already programmed schema $stack$ (implications are coded as disjunctions).}
\label{fig:compilation}
\end{figure}
When the observation $\tau$ includes observed actions, then the extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{ar(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ are included in the $\mathsf{apply_{\xi,\omega}}$ actions to validate that actions are applied, exclusively, in the same order as they appear in $\tau$.\\

>>>>

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a {\em classical planning problem} such that the semantics of the actions in $A$ is unknown because functions $\rho(s,a)$ and/or $\theta(s,a)$ are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite and non-empty {\em set of models} for the actions in $A$ such that, each model in $\mathcal{M}\in M$, defines a different function pair $\tup{\rho,\theta}$.
\item $\tau$ is an {\em observation} of the execution of a solution plan $\pi$ for $P$.
\end{itemize}

The {\em discrete probability distribution} $P(\mathcal{M}|\tau)$ expresses, for each model $\mathcal{M}\in M$, its probability of producing the observation $\tau$. The {\em naive Bayes classifier} assigns a model $\mathcal{M}\in M$ to the given observation $\tau$ wrt the following expression, $argmax_{\mathcal{M}\in M} P(\mathcal{M})\times P(\tau|\mathcal{M})$. With this regard, the solution to the {\em model recognition} task is the model $\mathcal{M}\in M$ that maximizes the previous expression.

According to the {\em Bayes} rule, the probability of an hypothesis $\mathcal{H}$, provided the observation $\mathcal{O}$, is given by the expression $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In the {\em model recognition} task, hypotheses are about the possible action models $\mathcal{M}\in M$ while the given observation is the partially observed plan execution $\tau$. The $P(\mathcal{M}|\tau)$ probability distribution can then be estimated in three steps:
\begin{enumerate}
\item Estimating the {\em a priori} probabilities, if they are not given as input. $P(\tau)$, that measures how surprising is the given observation and $P(\mathcal{M})$, that expresses if one model is a priori more likely than the others. 
\item Computing likelihoods $P(\tau|\mathcal{M})$.  Our approach is to estimate this value according to the {\em amount of edits} required by the model $\mathcal{M}$ to produce a plan $\pi$ such that:
\begin{enumerate}
\item $\pi$ solves $P$ and,
\item $\tau$ is an observation of the execution of $\pi$ on the classical planning problem $P$. 
\end{enumerate}
The {\em edit distance} is a similarity metric that is traditionally computed over strings or graphs and that has been proved succesful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work this similarity metric refers to the edition of classical planning models.
\item Applying the Bayes rule to obtain the normalized posterior probabilities. The $P(\mathcal{M}|\tau)$ probabilities must sum 1 for all the $\mathcal{M}\in M$.
\end{enumerate}

@@

\begin{figure}
\begin{tiny}
\begin{verbatim}
;;; Propositional encoding for inc-x(?v1 ?v2)
(pre_x-coord_v1_inc-x) (pre_next_v1_v2_inc-x) 
(pre_state-q0__inc-x)
(del_x-coord_v1_inc-x) (add_x-coord_v2_inc-x)

;;; Propositional encoding for dec-x(?v1 ?v2)
(pre_x-coord_v1_dec-x) (pre_next_v2_v1_dec-x) 
(pre_state-q1___dec-x)
(del_x-coord_v1_dec-x) (add_x-coord_v2_dec-x)

;;; Propositional encoding for inc-y-odd(?v1 ?v2)
(pre_y-coord_v1_inc-y-odd) (pre_next_v1_v2_inc-y-odd) 
(pre_state-q0__inc-y-odd)
(del_y-coord_v1_inc-y-odd) (del_state-q1__inc-y-odd)
(add_y-coord_v2_inc-y-odd) (add_state-q0__inc-y-odd)

;;; Propositional encoding for inc-y-even(?v1 ?v2)
(pre_y-coord_v1_inc-y-even) (pre_next_v1_v2_inc-y-even)
(pre_state-q0__inc-y-even)
(del_y-coord_v1_inc-y-even) (del_state-q0__inc-y-even)
(add_y-coord_v2_inc-y-even) (add_state-q1__inc-y-even)

;;; Propositional encoding for dec-y-odd(?v1 ?v2)
(pre_y-coord_v1_dec-y-odd) (pre_next_v2_v1_dec-y-odd)
(pre_state-q1__dec-y-odd)
(del_y-coord_v1_dec-y-odd) (del_state-q1__dec-y-odd)
(add_y-coord_v2_dec-y-odd) (add_state-q0__dec-y-odd)

;;; Propositional encoding for inc-y-even(v1 ?v2)
(pre_y-coord_v1_dec-y-even) (pre_next_v2_v1_dec-y-even)
(pre_state-q0__dec-y-even)
(del_y-coord_v1_dec-y-even) (del_state-q0__dec-y-even)
(add_y-coord_v2_dec-y-even) (add_state-q1__dec-y-even)
\end{verbatim}
\end{tiny}
 \caption{\small Propositional encoding for the six schema from Figure~\ref{fig:model-example}.}
\label{fig:encoding}
\end{figure}

@@

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action inc-x
  :parameters (?v1 ?v2)
  :precondition
    (and (or (not (pre_x-coord_v1_inc-x)) (x-coord ?v1))
         (or (not (pre_x-coord_v2_inc-x)) (x-coord ?v2))
         (or (not (pre_y-coord_v1_inc-x)) (x-coord ?v1))                       
         (or (not (pre_y-coord_v2_inc-x)) (x-coord ?v2))
         (or (not (pre_state-q0__inc-x)) (state-q0))
         (or (not (pre_state-q1__inc-x)) (state-q1)))
         (or (not (pre_next_v1_v1_inc-x)) (next ?v1 ?v1)))
         (or (not (pre_next_v1_v2_inc-x)) (next ?v1 ?v2)))
         (or (not (pre_next_v2_v1_inc-x)) (next ?v2 ?v1)))
         (or (not (pre_next_v2_v2_inc-x)) (next ?v2 ?v2)))         
    )
    :effect (and
       (when (del_x-coord_v1_inc-x) (not (x-coord ?v1)))
       (when (del_x-coord_v2_inc-x) (not (x-coord ?v2)))
       (when (del_y-coord_v1_inc-x) (not (x-coord ?v1)))
       (when (del_y-coord_v2_inc-x) (not (x-coord ?v2)))
       (when (del_state-q0__inc-x) (not (state-q0)))
       (when (del_state-q1__inc-x) (not (state-q1)))
       (when (del_next_v1_v1_inc-x) (not (next ?v1 ?v1)))
       (when (del_next_v1_v2_inc-x) (not (next ?v1 ?v2)))
       (when (del_next_v2_v1_inc-x) (not (next ?v2 ?v1)))
       (when (del_next_v2_v2_inc-x) (not (next ?v2 ?v2)))

       (when (add_x-coord_v1_inc-x) (x-coord ?v1))
       (when (add_x-coord_v2_inc-x) (x-coord ?v2))
       (when (add_y-coord_v1_inc-x) (x-coord ?v1))
       (when (add_y-coord_v2_inc-x) (x-coord ?v2))
       (when (add_state-q0__inc-x) (state-q0))
       (when (add_state-q1__inc-x) (state-q1))
       (when (add_next_v1_v1_inc-x) (next ?v1 ?v1))
       (when (add_next_v1_v2_inc-x) (next ?v1 ?v2))
       (when (add_next_v2_v1_inc-x) (next ?v2 ?v1))
       (when (add_next_v2_v2_inc-x) (next ?v2 ?v2)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Programmable version of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid.}
\label{fig:programmable}
\end{figure}

@@

For each action schema $\xi$, given the set of {\em state predicates} $\Psi$, then $F_{\xi}$ defines the subset of elements that can appear in the preconditions and effects of that action schema. For the actions schema defined in~\ref{fig:model-example} this set is the same and contains the following ten elements, {\small\tt\{x-coord($v_1$), x-coord($v_2$), y-coord($v_1$), y-coord($v_2$), state-q0(), state-q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

\begin{enumerate}
\setcounter{enumi}{2}
\item ${\mathcal C}$, a set of {\em syntactic constraints}. These constraints include the \strips\ constraints (we assume that $\eff^-(a)\subseteq \pre(a)$, $\eff^-(a)\cap \eff^+(a)=\emptyset$ and $\pre(a)\cap \eff^+(a)=\emptyset$) and domain-specific constraints. For instace, in a {\em robot navigation} domain like the modeled in Figure~\ref{fig:model-example}, predicates {\small\tt state-q0()} and {\small\tt state-q1()} are exclusive so they cannot hold a the same time.
\end{enumerate}

@@

Estimating the conditional probability $P(\tau|\mathcal{M})$ by mapping the {\em observation edit distance} into a $[0,1]$ likelihood, $1-\frac{\delta(\mathcal{M},\tau)}{\delta(\mathcal{M},*)}$.

@@

\item Actions for {\em applying} an edited operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:programmable} shows the PDDL encoding of the action for applying a programmed operator.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
&\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{edit}\}\rhd\{\neg mode_{edit}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

When the observation $\tau$ includes observed actions, then the extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{ar(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ are included in the $\mathsf{apply_{\xi,\omega}}$ actions to validate that actions are applied, exclusively, in the same order as they appear in $\tau$.\\
